---
title: "HW2-Data Cleaning"
author: "Eva Newby"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
## Instructions - 3 steps
### 1. Clean the Water_cover column to transform it into the correct data type and respect expectations for a percentage

```{r}
#| message: false  

# Load packages
library(tidyverse)

# load file names
datadir_raw <- file.path("data","raw")

datadir_processed <- file.path("data", "processed/")

species_file <- "ASDN_Daily_species.csv"

snowsurvey_file <- "ASDN_Snow_survey.csv"

# Import the species
snowsurvey_csv <- read_csv(file.path(datadir_raw, snowsurvey_file))
```

```{r}
# View the data to see what needs cleaning
glimpse(snowsurvey_csv)

# Count the occurrences of each unique value in the water column
snowsurvey_csv %>%
  count(Water_cover)
```

There are several non-numeric columns that need to be addressed; namely, "-", and ".". 

```{r}

```

```{r}

```

```{r}
# View the datatype for the water column
class(snowsurvey_csv$Water_cover)
```

This returns a 'character', meaning that we need to change the type to numeric. Now that we've addressed all our non-numeric columns, let's change the data type. 
```{r}
# Change data type from character to numeric

```

```{r}

```

### 2. Clean the Land_cover column to transform it into the correct data type and respect expectations for a percentage

```{r}

```

### 3. Use the relationship between the three cover columns (Snow, Water, Land) to infer missing values where possible and recompute the Total_cover column as needed

```{r}

```

